---
title: "COVID-19 US Confirmed Cases and Deaths Data Analysis"
# author: 
date: "2023-12-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(ggplot2)
library(dplyr)
library(viridis)
library(sf)
library(tidyverse)
library(usmap)
library(maps)
```

## Overview

This report is an analysis of multiple COVID data sets available from John's Hopkins github page. 
The data sets are as follows:

* [US Confirmed Cases](https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv): Time series data of confirmed COVID-19 cases in the US.
* [US Deaths](https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv): Time series data of COVID-19-related deaths in the US.

I will retrieve population data for each US state:

* [US State Population](https://gist.githubusercontent.com/bradoyler/0fd473541083cfa9ea6b5da57b08461c/raw/fa5f59ff1ce7ad9ff792e223b9ac05c564b7c0fe/us-state-populations.csv)

I will create variables for the __US Confirmed Cases__ and __US Deaths__ datasets:
```{r Data Set Varialbles, echo=TRUE}
# US Confirmed cases
us_confirmed <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")

# US Deaths
us_deaths <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")

usStatePop <- read.csv('https://raw.githubusercontent.com/ColoradoGus/DTSA5301/main/NST-EST2022-POPCHG2020_2022.csv')
usStatePop <- select(usStatePop, NAME, POPESTIMATE2022)
```

Using this data set, I will do some high-level analysis of the data set.
First, I'm going to focus on US confirmed COVID-19 cases by analyzing the dataframe called **us_confirmed**.

## US Confirmed Cases Analysis

### DataFrame Overview

In __us_confirmed__, there are `r nrow(us_confirmed)` rows and `r ncol(us_confirmed)` columns. Given that there are over 1000 column headers. This dataset counts cases from 1/22/20-3/9/23. Each date from 01/22/20-3/9/23 is a column.
The dataset shows cumulative confirmed cases. Therefore, the value in column 3/9/23 shows cases for the full two year period. Below you can see the header for a few key columns: "Latitude", "Longitude", "Province_State", and "X3.9.23" (the last date in the file).

```{r Viewing the US Confirmed Dataset, echo=TRUE}
# head of us_confirmed
head(us_confirmed[, c("Lat", "Long_", "Province_State", "X3.9.23")])
```

### Creating Daily Cases Total DataFrame
Let's look at daily confirmed cases as a line chart. First we need to create daily totals of cases by date.

```{r Creating a daily totals dataframe, echo=TRUE}
# Sum the cases for each date
daily_totals <- colSums(us_confirmed[, 12:(ncol(us_confirmed))])

# Convert the date fields
dates <- sub("^X", "", names(daily_totals))
formatted_dates <- as.Date(dates, format = "%m.%d.%y")

# Create the dataframe for plotting daily totals
daily_data <- data.frame(
  Date = formatted_dates,
  Cases = as.numeric(daily_totals)
)

head(daily_data)
```

### Plotting Daily Totals

Cumulative confirmed cases reached a total of `r format(sum(us_confirmed$X3.9.23), big.mark = ",", scientific = FALSE)` cases. Remember, people can get COVID-19 multiple times. It is hard to know how many unique people had COVID-19.
```{r Plotting Daily Confirmed Cases Totals, echo=TRUE}
ggplot(daily_data, aes(x = Date, y = Cases)) +
  geom_line(color = "blue") +
  labs(title = "Cumulative Daily Confirmed COVID-19 Cases in the US",
       x = "Date",
       y = "Number of Cumulative Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The graph makes it clear that COVID cases spiked in the winter.

## US Confirmed Deaths

### Reviewing Confirmed Deaths DataFrame
Now I'd like to look at confirmed deaths. There are `r format(sum(us_deaths$X3.9.23), big.mark = ",", scientific = FALSE)` confirmed deaths in the data set. 

```{r Summarizing death totals, echo=TRUE}
# The head of the us_deaths dataset.
head(us_deaths[, c("Lat", "Long_", "Province_State", "X3.9.23")])
```

### Creating a Cumulative Daily COVID Deaths DataFrame

In order to plot the cumulative daily deaths, I want to create a new dataframe with one row per date and the cumulative deaths for that date. The tail of the dataframe is shown below. 
```{r Creating Daily Death DataFrame, echo=TRUE}
# Sum the cases for each date
daily_death_totals <- colSums(us_deaths[, 13:(ncol(us_deaths))])

# Convert the data into a format suitable for ggplot2
dates <- sub("^X", "", names(daily_death_totals))
formatted_dates <- as.Date(dates, format = "%m.%d.%y")

# Create the dataframe for plotting
daily_death_data <- data.frame(
  Date = formatted_dates,
  Cases = as.numeric(daily_death_totals)
)

tail(daily_death_totals)
```

### Plotting Daily Deaths

I will plot cumulative daily confirmed deaths, in the same way I plotted daily confirmed cases.
```{r Plotting Daily Confirmed Death Totals, echo=TRUE}
ggplot(daily_death_data, aes(x = Date, y = Cases)) +
  geom_line(color = "red") +
  labs(title = "Cumulative Daily COVID-19 Deaths in the US",
       x = "Date",
       y = "Number of Cumulative Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Visualizing Cases and Deaths in a Single Plot

I want to visualize cases and deaths in the same plot. First, I will create a merged dataframe of cases and deaths.

### Creating a Merged DataFrame

I will create a merged dataframe and calculate the log of each column as well.
```{r Creating a merged dataset, echo=TRUE}
# Sum the cases for each date
mergedDailyData = merge(daily_data, daily_death_data, by="Date")
names(mergedDailyData)[names(mergedDailyData) == "Cases.y"] <- "Deaths"
names(mergedDailyData)[names(mergedDailyData) == "Cases.x"] <- "Cases"

mergedDailyData$logCases = log(mergedDailyData$Cases, base = 10)
mergedDailyData$logDeaths = log(mergedDailyData$Deaths, base = 10)
mergedDailyData$deathsPerCase = mergedDailyData$Deaths/mergedDailyData$Cases
last_value <- round(tail(mergedDailyData$deathsPerCase, 1),2)

head(mergedDailyData)
```
### Plotting the Log10 of Confirmed Cases and Deaths

Since the scale of cases and deaths differ by two orders of magnitude, I want to plot the cases and deaths on a log scale. In this plot, we can better see a comparision between the Cumulative cases and Deaths over time.
```{r Plotting Daily Confirmed Cases and Death Totals, echo=TRUE}
ggplot(mergedDailyData, aes(x = Date)) +
  geom_line(aes(y = logCases, color = "Cases")) +
  geom_line(aes(y = logDeaths, color = "Deaths")) +
  labs(title = "Cases and Deaths Over Time", x = "Date", y = "Log10 Count") +
  scale_color_manual(values = c("Cases" = "blue", "Deaths" = "red"))

```

## Modeling and Visualizing Case Fatality Rate

In the merged DataFrame I created, I calculated the fatality rate. The case fatality rate appears to be around `r last_value`. I'd like to visualize the case fatality rate over the data's date range. 

```{r Modeling Fatality Rate, echo=TRUE}
ggplot(mergedDailyData, aes(x = Date, y = deathsPerCase)) +
  geom_line(color = "red") +
  labs(title = "Cumulative Case Fatality Rate in the US",
       x = "Date",
       y = "Case Fatality Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 0.1)

```

## Modeling and Visualizing Per Capita Deaths By State

I'd like to model and visualize deaths per capita by state. I'll first create a dataframe with total deaths per state.

```{r Deaths By State}
selected_data <- us_deaths %>% select(Province_State, `X3.9.23`)
total_cases_by_state <- (selected_data %>% 
                        group_by(Province_State) %>% 
                        summarise(Total_Cases = sum(`X3.9.23`, na.rm = TRUE)))

usStatePop$stateLower = tolower(usStatePop$NAME)
head(total_cases_by_state)
```

Then I will get a dataframe of states maps and population. I will append the death totals to it and create a per-capita data column. 

```{r  Creating a Death Per Capita by State DataFrame}

states_map <- st_as_sf(maps::map('state', plot = FALSE, fill = TRUE))
total_cases_by_state$Province_State <- tolower(total_cases_by_state$Province_State)
total_cases_by_state$log_totals <- log(total_cases_by_state$Total_Cases, base = 10)
merged_data <- merge(states_map, total_cases_by_state, by.x = 'ID', by.y = 'Province_State')
merged_data <- merge(merged_data, usStatePop, by.x = 'ID', by.y = 'stateLower')
merged_data$per_capita_total = merged_data$Total_Cases/merged_data$POPESTIMATE2022 *100000
```

Then I will create a [choropleth map](https://en.wikipedia.org/wiki/Choropleth_map) of the data.
```{r Creating a Death Per Capita Choropleth}
ggplot(merged_data) +
  geom_sf(aes(fill = per_capita_total)) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "COVID-19 Per-Capita Total Deaths by State as of 3/9/23",
       fill = "Per Capita Cases per 100,000") +
  theme_minimal()
```

Based on the choropleth above, it looks like the midwest, southern, and sun belt states had some of the highest death rates per capita.

## Conclusion and Biases

### Overall 
To summarize, I analyzed the US Covid Confirmed Cases and Confirmed Deaths data sets.
I visualized the log Case and Death totals and modeled the case fatality rate. I also modeled the total deaths per capita by state. Overall, this was an interesting data set on which to perform basic analysis. Also, it highlights how the modern world can track so much health informaton. This data set is just the tip of the iceberg! 

### Biases

There can be a number of biases in the data sets. For deaths, it is possible that Covid could be over-attributed to the cause of death. However it is possible early on that the cases were under reported before teating got started. Additionally, it can be hard to interpret the case fatality rate because the vaccine became available during this time interval. 