---
title: "NYPD Shooting Analysis"
date: "2023-10-15"
output: html_document
---

```{r setup, include}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(ggplot2)
library(dplyr)
library(viridis)
```

## Overview

This report is an analysis of the [NYPD shooting data set](https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic/resource/c564b578-fd8a-4005-8365-34150d306cc4).
This data set a list of every shooting incident that occurred NYC from 2006 to 2022.

I will:

* Count the data rows, list the column headers, and discuss the data fields I will keep for analysis
* Create a boxplot by month of shootings
* Create a scatter plot of shootings by longitude and latitude
* Evaluate per capita shootings
* Discuss bias that could exist in the data set

### Data Intake and Cleanup

I will create a dataframe called **nypdShooting** using the provided link. 

```{r NYPD Shooting Data}
nypdShooting <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")

```

There are `r nrow(nypdShooting)` rows and `r ncol(nypdShooting)` columns. The column headers are:

`r paste0("* ", names(nypdShooting), collapse = "\n")`

#### Data Tiding

I'm going to remove columns with limited or redundant data and keep only the columns listed in the code blocks below.

```{r Cleaning NYPD Data Set}
colsKeep <- c("OCCUR_DATE", "OCCUR_TIME", "BORO", "PRECINCT", "STATISTICAL_MURDER_FLAG", 
                     "VIC_SEX", "VIC_RACE", "X_COORD_CD", "Y_COORD_CD")

nypdShooting <- nypdShooting[, colsKeep]
```

I will use this smaller dataframe for my remaining analysis.
Looking at this smaller dataframe, all columns have data (no null values).
Therefore, I will consider this data set complete for my purposes.

```{r Checking for Null Values}
nullCounts <- colSums(is.na(nypdShooting))
nullCounts
```

### Data Vizualization

#### Month of Year Analysis

I want to see how shootings change throughout the year, if at all. I will visualize this in two ways: 

1) A heatmap of incidents by month and year.
2) A boxplot of incidents by month

To do this I will first need to summarize the data by month and year:

```{r Summarizing Incident Data}
nypdShooting$Year <- year(as.Date(nypdShooting$OCCUR_DATE, format="%m/%d/%Y"))
nypdShooting$Month <- month(as.Date(nypdShooting$OCCUR_DATE, format="%m/%d/%Y"), label = TRUE)

incidentCounts <- nypdShooting %>%
  group_by(Year, Month) %>%
  summarise(Incidents = n())
```

Based on the graphics below, there are more shootings in the summer months. 
Additionally, shootings during the summer of 2020 were high compared to other years.

```{r HeatMap of Shooting Incident Counts by Month and Year}
ggplot(incidentCounts, aes(x = Month, y = as.factor(Year), fill = Incidents)) +
  geom_tile() +
  scale_fill_gradientn(colors = viridis::viridis(8)) +
  labs(x = "Month", y = "Year", fill = "Incident Count", title = "Heatmap of Incidents by Month and Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r Shooting Incident Month of Year Boxplot}
ggplot(incidentCounts, aes(x = Month, y = Incidents)) +
  geom_boxplot() +
  labs(x = "Month", y = "Incident Count", title = "Distribution of Incidents by Month") +
  theme_minimal()
```

```{r Shooting Incident Month by Time of Day Boxplot}
#Trying something
nypdShooting$Hour <- as.numeric(substr(nypdShooting$OCCUR_TIME, 1, 2))

incidentHoursCounts <- nypdShooting %>%
  group_by(Month,Hour) %>%
  summarise(Incidents = n())

# ggplot(incidentHoursCounts, aes(x = Month, y = as.factor(Hour), fill = Incidents)) +
#   geom_tile() +
#   scale_fill_gradientn(colors = viridis::viridis(8)) +
#   labs(x = "Month", y = "Hour", fill = "Incident Count", title = "Heatmap of Incidents by Month and Year") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(incidentHoursCounts, aes(x = Hour, y = as.factor(Month), fill = Incidents)) +
  geom_tile() +
  scale_fill_gradientn(colors = viridis::viridis(8)) +
  labs(x = "Hour of Day", y = "Month", 
       fill = "Incident Count", 
       title = "Heatmap of Incidents by Month and Hour",
      ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ggplot(nypdShooting, aes(x = Month, y = OCCUR_TIME)) +
#   geom_point(alpha = 0.5, size = 0.7, color = "blue") +
#   labs(x = "Month", y = "Occurrence Time", title = "Scatter Plot of Occurrence Time") +
#   theme_minimal()

head(nypdShooting)
```


#### Geospatial Analysis (Scatter Plot)

I want to see where most shootings occur in NYC. I will simply create a scatter of longitude and latitude shooting events.
From this analysis it looks like the Bronx and Brooklyn have the most number of shootings.

```{r Shooting Incidents Longitude-Latitude Scatter Plot}

ggplot(nypdShooting, aes(x = X_COORD_CD, y = Y_COORD_CD)) +
  geom_point(alpha = 0.5, size = 0.7, color = "blue") +
  labs(x = "Longitude", y = "Latitude", title = "Scatter Plot of Shooting Incidents in NYC") +
  theme_minimal()

```


### Data Modeling

#### Investigate Per Capita Shootings

I'm going to see if how the per-capita shootings compare per borough. To do this I will get the populations per borough.
I will use [wikipedia](https://en.wikipedia.org/wiki/Boroughs_of_New_York_City) for the population information. The Bronx has 36 shootings per 100,000 residents, double that of Manhattan which has a higher population. 


```{r Per Capita Summary, echo}

data2022 <- nypdShooting %>%
  filter(Year == 2022)

boroughPop <- data.frame(
  BORO = c("BRONX", "BROOKLYN", "MANHATTAN", "QUEENS", "STATEN ISLAND"),
  Population = c(1472654, 2736074, 1694251, 2405464, 495747)
)

boroughShootings <- aggregate(OCCUR_DATE ~ BORO, data = data2022, FUN = length)
colnames(boroughShootings)[2] <- "Shootings"

# Merge the two data frames by borough
boroughPerCapita <- merge(boroughPop, boroughShootings, by = "BORO")

# Calculate per capita shootings per 100,000 residents
boroughPerCapita$PerCapitaShootings <- (boroughPerCapita$Shootings / boroughPerCapita$Population)*100000

print(boroughPerCapita)
```

### Conclusion

The raw data shows over 27000 shootings. It is possible that this data is biased. A few reasons are:

* Shootings could be under-reported in some areas 
* Higher police activity in some areas could skew reporting higher
* Changing police activity over years could bias the data for specific years

However, overall this data set is useful for giving an overview of crime activity in NYC and I imagine it is useful for policy makers and the police force.

